---
title: "Visual analytics of hotel bookings data"
author: "Julià Minguillón"
date: "April 2025"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

NOTE: this tutorial uses R + RStudio + some R packages to show the potential
of using data visualization for inspecting and analyzing a data set. We strongly
recommend you to explore the following links:

1) RStudio: https://posit.co/downloads/
2) ggplot2: https://ggplot2.tidyverse.org/
3) extensiones: https://exts.ggplot2.tidyverse.org/gallery/

## Load packages

```{r packages}
library("ggmosaic")
library("ggplot2")
library("fitdistrplus")
library("MASS")
library("survival")
library("ggstatsplot")
library("tidyverse")
```

## Data loading and dimensions (N x M)

We read the dataset in CSV format, with 119,390 rows y 32 columns:

```{r load data}
x=read.csv("hotel_bookings.csv", stringsAsFactors = T)
dim(x)
```
## Data cleansing

First, we'll inspect the data using the summary() function included in R.  You 
can find an explanation of each variable in the article that describes this 
dataset in detail, although the variable names are pretty much self-explanatory:

```{r summary, echo=FALSE}
summary(x)
```

# Numerical variables

Some unexpected (outliers?) values for several variables can be observed.
For instance:

1) A maximum of 55 in 'adults'
2) A maximum of 10 in 'children' (including also missing values)
3) A maximum of 10 in 'babies'
4) Negative values in the average daily rate ('adr') or or very high

Let's visualize the histogram of the variable 'adults', with at least 55 
breaks in the histogram, using the function hist() in R:

```{r hist_adults}
str(x$adults)
hist(x$adults,breaks=55)
sum(is.na(x$adults)) # check if there are NA values
```

It can be observed that the histogram shows no bars around the value 55, 
given that this is a very large set and probably it's only one or a few cases. 
In these cases, to analyze the extreme values of a variable, the values of the 
variable in question can be represented graphically as follows, ordering and 
plotting the data (if they are numerical, as in this case):

```{r plot_adults}
plot(sort(x$adults))
grid()
```
The 'Index' represents the position of the element once it's sorted, but we're 
more interested in the Y axis, as we can see that some elements have values 
of 10 or higher. Since this is an integer variable with a limited set of 
possible values, we can use table() to visualize them:

```{r table_adults}
table(x$adults)
```
As you can see, there's one reservation for 10 adults, two for 20 adults, 
and so on, up to one for 55 adults! Without going into further detail, we'll 
remove all rows with reservations for 10 or more adults:

```{r clean_adults}
x=x[x$adults<10,]
```

EXERCISE: Repeat this process with variables 'children' and 'babies'. Try also
to change the threshold to less than 5 instead of 10. 

<font color="blue">
**2)** A maximum of 5 in 'children' (including also missing values)

```{r hist_children}
hist(x$children,breaks=55)
x[is.na(x$children), ] # check if there are NA values
```

```{r plot_children}
plot(sort(x$children))
grid()
```

```{r table_children}
table(x$children)
```

As you can see, there's one reservation for 10 children. I will 
remove all rows with reservations for 5 or more children:

```{r clean_children}
x = x[!is.na(x$children) & x$children < 5, ]
#x=x[x$children<5,]
```

**3)** A maximum of 5 in 'babies'

```{r hist_babies}
hist(x$babies,breaks=55)
x[is.na(x$babies), ] # check if there are NA values
```

```{r plot_babies}
plot(sort(x$babies))
grid()
```

```{r table_babies}
table(x$babies)
```

As you can see, there's one reservation for 10 babies and another reservation for 9 babies. I will 
remove all rows with reservations for 5 or more babies:

```{r clean_babies}
x = x[x$babies < 5,]
```

</font>

The histogram of the 'adr' variable (average daily rate) presents the same 
problem as the 'adults' variable, so we will simply create a graph with the 
ordered values again:

```{r plot_adr}
plot(sort(x$adr))
grid()
```
In this case, we observe that only one value is significantly higher than the 
rest. We consider it an outlier and eliminate it, as well as the negative 
values which have no a clear explanation, although we keep the 0 values:

```{r clean_adr}
x=x[x$adr>=0 & x$adr<1000,]
```

The histogram now provides us with some relevant information. We draw it using
the ggplot2 package, which offers many more options than hist():

```{r hist_adr}
ggplot(data=x, aes(x=adr)) + 
  geom_histogram(bins=55, colour="black", fill = "lightgray") +
  theme_light()
```
EXERCISE: improve the graph to make axis, title, etc. more adequate.

```{r hist_adr_evo}

ggplot(data = x, aes(x = adr)) + 
  geom_histogram(bins = 55, color = "black", fill = "lightblue", alpha = 0.7) +
  labs(
    title = "Distribution of Average Daily Rate (ADR)",
    x = "Average Daily Rate (ADR)",
    y = "Count of Observations"
  ) +
  theme_light(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title.x = element_text(face = "bold"),
    axis.title.y = element_text(face = "bold")
  ) +
  scale_x_continuous(labels = scales::dollar_format(prefix = "€"))
```

We can see that there is a set of approximately 2,000 zero values, which could
be analyzed separately, for example. There are R packages that help us estimate
this distribution and the parameters that determine it visually, such as the 
fitdistrplus package, which provides the descdist() function (caution, slow!):

```{r descdist}
require(fitdistrplus)
descdist(x$adr,boot=1000)
```
As you can see, the real data (observations, a colored dot) and the simulated
data (in other color) approximate what a lognormal distribution might look like. 
However, to experiment with the cleanest possible data set, we will:

1) remove 0-day stays
2) remove 0-cost stays
3) remove stays with no guests
4) replace the NAs in the children variable with 0

```{r data_cleansing}
x[is.na(x$children),'children']=0
x=x[x$adr>0 & 
    (x$stays_in_week_nights+x$stays_in_weekend_nights)>0 & 
    (x$adults+x$children+x$babies)>0 & 
    !is.na(x$children),]

x[is.na(x$children), ]
```

## Categorical variables

For categorical variables, the summary() function gives us a first idea of the
possible values each can take. For example, in the original set (before removing
outliers), there are 79,330 reservations at a city hotel (Lisbon) and 40,060 at
a resort (Algarve). We can ask ourselves whether the cost distribution is the
same for both groups, either by using the appropriate statistical test or simply
by comparing histograms, in this case using the ggplot2 package, which is much
more powerful for creating all kinds of graphs:

```{r hist_adr_tipo}
# require(ggplot2)
ggplot(data=x, aes(x=adr, fill=hotel)) + 
  geom_histogram(bins=50, colour="black") +
  theme_light()
```
It can be seen that the most common prices in Lisbon (city hotels) are slightly
to the right of the most common prices in the Algarve (resort hotels), although
the highest prices in Lisbon decrease more rapidly than in the Algarve. By using
a violin plot, we can see more detail, especially if we also show the typical
quartiles of a box plot:

```{r violin_adr_tipo}
ggplot(data=x, aes(x=hotel, y=adr, fill=hotel)) + 
  geom_violin() + geom_boxplot(width=.1, outliers = F) +
  coord_flip() + 
  theme_light()
```
There is an R package called ggstatsplot that has specific functions for each
type of graph, including appropriate statistical tests to determine if there
are differences between groups:

```{r ggstatsplot}
# require(ggstatsplot)
ggbetweenstats(data=x, x=hotel, y=adr)
```
Another interesting variable is the hotel guests' origin ('country'). The
problem is that this variable has many different values (178), so we should
focus on the countries with the most tourists, also showing whether they choose
a city hotel or a resort:

```{r country}
# require(tidyverse)
# countries with at least 100 bookings
xx = x %>% group_by(country) %>% mutate(pais=n()) %>% filter(pais>=100)
xx$country=factor(xx$country)
ggplot(data=xx, aes(x=reorder(country, -pais))) + 
  geom_bar(stat="count", aes(fill=hotel)) +
  theme_light() + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```
Obviously, Portugal (PRT) ranks first, followed by neighboring countries such
as Great Britain, France, and Spain. Visitors from Great Britain and Ireland
are most likely to choose a resort, while those from France, Germany, and Italy
primarily visit Lisbon.

<font color="blue">

EXERCISE: Are there differences between residents of Portugal and the rest?
Yes, the number of people that go to both city and resort are much higher than the rest, obviusly beacause they are local residents.

</font>

Another interesting variable is 'is_canceled', which indicates whether a 
reservation was canceled or not (37.0% of the time). We can observe the 
relationship between two categorical variables using a mosaic chart:

```{r mosaic_hotel_is_canceled}
# require(ggmosaic)
x$is_canceled=as.factor(x$is_canceled)
ggplot(data=x) + 
  geom_mosaic(aes(x=product(is_canceled, hotel), fill=hotel)) +
  theme_light() 
```
It can be seen that the cancellation rate (denoted by 1 on the Y-axis) at a 
resort is lower than that of a hotel in Lisbon. On the X-axis, the relative
size of each column also corresponds to the proportion of each hotel type. 
It is important not to consider the Y-axis labels (0/1) as the actual numerical
cancellation rate, as this can be misleading.

<font color="blue">

EXERCISE: which other type of graph could be used to represent this data?


```{r otherplot_hotel_is_canceled}

# Copia del dataset original
dfx1 <- x

# Asegurar factores legibles
dfx1$is_canceled <- factor(dfx1$is_canceled, levels = c(0, 1), labels = c("No Cancelado", "Cancelado"))

# Calcular proporciones y etiquetas
dfx1_pie <- dfx1 %>%
  count(hotel, is_canceled) %>%
  group_by(hotel) %>%
  mutate(
    prop = n / sum(n),
    pct_label = paste0(round(prop * 100, 1), "%"),
    ypos = cumsum(prop) - prop / 2  # posición vertical dentro del segmento
  )

# Gráfico tipo pie con etiquetas
ggplot(dfx1_pie, aes(x = "", y = prop, fill = is_canceled)) +
  geom_col(width = 1, color = "white") +
  coord_polar("y") +
  facet_wrap(~ hotel) +
  geom_text(aes(y = ypos, label = pct_label), color = "white", size = 4, fontface = "bold") +
  labs(
    title = "Proporción de cancelaciones por tipo de hotel",
    fill = "Estado de reserva"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(face = "bold")
  )


```

</font>

In the case of cancellation by country for the countries with more tourists:

```{r mosaic_country_is_canceled}
# at least 1000 bookings
xx = x %>% group_by(country) %>% mutate(pais=n()) %>% filter(pais>=1000)
xx$country=factor(xx$country)
ggplot(data=xx) + 
  geom_mosaic(aes(x=product(is_canceled, country), fill=country)) +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```
It can be seen that the cancellation rate is much higher for local tourists 
(from Portugal, PRT), while it is much lower for the rest of the countries.
However, this graph is not easy to read; in this case, there is no order of 
either the countries or the percentage of cancellations.

<font color="blue">

EXERCISE: Improve the previous graph to make it more understandable and 
consider whether it is possible to visualize the relationships between three
or more categorical variables.

**1. Use readable labels for is_canceled**
Convert it to a labeled factor: "Canceled" vs "Not Canceled".

**2. Control the number of countries shown**
Even with ≥1000 bookings, many countries can still overwhelm the plot. Consider showing only the top 10–15 by count.

**3. Enhance color palette**
Using a perceptually uniform or qualitative palette helps distinguish countries better.

```{r mosaic2_country_is_canceled}
# Copy and prep
dfx2 <- x %>%
  mutate(is_canceled = factor(is_canceled, levels = c(0,1), labels = c("Not Canceled", "Canceled"))) %>%
  group_by(country) %>%
  mutate(pais = n()) %>%
  filter(pais >= 1000) %>%
  ungroup()

# Keep only top 10 countries
top_countries <- dfx2 %>%
  count(country, sort = TRUE) %>%
  top_n(10) %>%
  pull(country)

dfx2 <- dfx2 %>% filter(country %in% top_countries)

# Transforms Country to factor
dfx2$country=factor(dfx2$country)

# Plot
ggplot(data = dfx2) + 
  geom_mosaic(aes(x = product(is_canceled, country), fill = country)) +
  theme_light() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    legend.position = "right"
  ) +
  labs(
    title = "Cancellation Rate by Country (Top 10, ≥1000 bookings)",
    x = "Cancellation Status and Country",
    fill = "Country"
  )
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) 
```

**4. Alternative: Grouped Bar Plot (simpler & cleaner)**
Mosaic plots are useful but get messy with more categories. Bar plots are often easier to interpret.

```{r barplot_country_is_canceled}
# Copy and prep
dfx2 <- x %>%
  mutate(is_canceled = factor(is_canceled, levels = c(0,1), labels = c("Not Canceled", "Canceled"))) %>%
  group_by(country) %>%
  mutate(pais = n()) %>%
  filter(pais >= 1000) %>%
  ungroup()

# Keep only top 10 countries
top_countries <- dfx2 %>%
  count(country, sort = TRUE) %>%
  top_n(10) %>%
  pull(country)

dfx2 <- dfx2 %>% filter(country %in% top_countries)

df_summary <- dfx2 %>%
  count(country, is_canceled) %>%
  group_by(country) %>%
  mutate(prop = n / sum(n))

ggplot(df_summary, aes(x = reorder(country, -prop), y = prop, fill = is_canceled)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Proportion of Cancellations by Country", y = "Proportion", x = "Country") +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r barplot2_country_is_canceled}

ggplot(dfx2, aes(x = is_canceled, fill = is_canceled)) +
  geom_bar() +
  facet_wrap(~ country, scales = "free_y") +
  labs(
    title = "Cantidad de cancelaciones por país (Top 15)",
    x = "Estado de cancelación",
    y = "Número de reservas"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))

```


</font>

Finally, let's analyze the behavior of reservations relative to the arrival
date. First, using the R lubridate package (a marvel for manipulating date and
time data), we'll create a 'day' variable to determine the day of the week the
hotel was checked in and analyze how many reservations there were each day:

```{r dia}
# require(lubridate)
x$dia=as_date(paste0(x$arrival_date_year,'-',x$arrival_date_month,'-',x$arrival_date_day_of_month))
ggplot(data=x,aes(x=dia,group=arrival_date_year,color=as.factor(arrival_date_year))) + 
  geom_bar() + scale_color_manual(values=c("2015"="red","2016"="green","2017"="blue")) + 
  theme_light() + 
  theme(legend.position='none') 
```

<font color="blue">

EXERCISE: Improve and split the above graph by hotel type or country of origin.

**1. By hotel**

We'll also use a line plot (geom_line(stat="count")) instead of bars for smoother time series visualization and include a legend.
```{r dia_by_hotel}

# Copia del dataset
dfx3 <- x

# Crear columna de fecha
dfx3$dia <- as_date(paste0(x$arrival_date_year,'-',x$arrival_date_month,'-',x$arrival_date_day_of_month))

# Agregar número de reservas por día y año
dfx3_plot <- dfx3 %>%
  group_by(hotel, dia, arrival_date_year) %>%
  summarise(n = n(), .groups = "drop") %>%
  mutate(year = as.factor(arrival_date_year))


# Graficar con líneas
ggplot(dfx3_plot, aes(x = dia, y = n, color = year)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~ hotel, ncol = 1) +
  scale_color_manual(values = c("2015" = "red", "2016" = "green", "2017" = "blue")) +
  labs(
    title = "Daily Booking Trends by Hotel Type",
    x = "Date of Arrival",
    y = "Number of Bookings",
    color = "Year"
  ) +
  theme_light() +
  theme(
    legend.position = "top",
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

**2. By country**
```{r dia_by_country}

# Filter top 5 countries
top_countries <- dfx3 %>%
  count(country, sort = TRUE) %>%
  top_n(5) %>%
  pull(country)

dfx3 <- dfx3 %>% filter(country %in% top_countries)

# Transforms Country to factor
dfx3$country=factor(dfx3$country)

# Plot
ggplot(data = dfx3, aes(x = dia, group = arrival_date_year, color = as.factor(arrival_date_year))) + 
  geom_line(stat = "count", size = 0.7) +
  scale_color_manual(values = c("2015" = "red", "2016" = "green", "2017" = "blue")) +
  facet_wrap(~ country, ncol = 1) +
  labs(
    title = "Daily Booking Trends by Country of Origin (Top 5)",
    x = "Date of Arrival",
    y = "Number of Bookings",
    color = "Year"
  ) +
  theme_light() +
  theme(
    legend.position = "top",
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


```

</font>

As described in the article, the data covers the period from July 1, 2015, to
August 31, 2017. Some peaks can be observed that might be interesting to explain
(what happened those days, i.e. 2015-12-05?). You can check Google Trends to
get some insights:

https://trends.google.es/trends/explore?date=2015-01-01%202017-12-31&q=lisboa,algarve&hl=es

```{r max_dia}
max(table(x$dia))
which.max(table(x$dia))
```
With the computed day 'dia', along with the variables 'stays_in_week' and 
'weekend_nights', we can try to manually categorize the trip type according
to the following criteria (this is arbitrary, clearly improvable):

1) if 'stays_in_weekend_nights' is zero => work trip
2) if 'stays_in_week_nights' is zero or one and in this case the entry is on
   Friday => weekend
3) if 'stays_in_week_nights' is five and 'stays_in_weekend_nights' is three 
  (that is, from Saturday or Sunday to Saturday or Sunday) 
   => week holiday package
4) if 'stays_in_weekend_nights' is one or two and 'stays_in_week_days' is five
   or less => work + rest
5) the rest of combinations => holidays

```{r tipo_visita}
# require(lubridate)
x$tipo=ifelse(x$stays_in_weekend_nights==0, "work",
       ifelse(x$stays_in_week_nights==0, "weekend",
       ifelse(x$stays_in_week_nights==1 & wday(x$dia)==6, "weekend",
       ifelse(x$stays_in_week_nights==5 & 
              (x$stays_in_weekend_nights==3 |
               x$stays_in_weekend_nights==4), "package",
       ifelse(x$stays_in_week_nights<=5 & 
              x$stays_in_weekend_nights<3, "work+rest",
       "rest")))))
```

One way to refine this classification would be to look at the number of adults,
children, and infants to decide whether it is a business traveler or a family.
The possibilities are endless: you can enrich the dataset with geographic data
(distance between countries), demographic data, economic data (per capita
income), weather data (in both Portugal and the country of origin), etc. 

<font color="blue">

EXERCISE: You must explore such enriched dataset and, in this process of 
exploration, decide what story you want to tell about it. Some ideas:

**1)** do tourists from different countries travel in different dates?

```{r dif_country_dif_area}

# Asegurar que 'dia' es tipo fecha
x$dia <- as.Date(x$dia)

# Obtener los 10 países con más turistas
top_5_countries <- x %>%
  count(country, name = "total") %>%
  arrange(desc(total)) %>%
  slice_head(n = 5) %>%
  pull(country)

# Filtrar dataset
x_top <- x %>%
  filter(country %in% top_5_countries)

# Agrupar por día y país
grouped_top <- x_top %>%
  group_by(dia, country) %>%
  summarise(count = n(), .groups = "drop")

# Graficar líneas suavizadas
ggplot(grouped_top, aes(x = dia, y = count, color = country)) +
  geom_smooth(se = FALSE, method = "loess", span = 0.3, linewidth = 1.2) +
  labs(
    title = "Tendencia de turistas por país (Top 5)",
    x = "Fecha",
    y = "Cantidad de turistas (tendencia suavizada)",
    color = "País"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

**2)** differences in cancellations among groups (countries, type of stay, ...)

% of cancellations per country


```{r dif_cancellation_country}

# Top 10 países con más reservas
top_countries <- x %>%
  count(country, sort = TRUE) %>%
  slice_head(n = 10) %>%
  pull(country)

# Filtrar
x_top <- x %>%
  filter(country %in% top_countries)

x_top <- x_top %>%
  mutate(is_canceled = as.numeric(as.character(is_canceled)))

# Calcular proporción de cancelaciones
cancel_by_country <- x_top %>%
  group_by(country) %>%
  summarise(
    cancel_rate = mean(is_canceled),
    total = n()
  )

# Graficar
ggplot(cancel_by_country, aes(x = reorder(country, -cancel_rate), y = cancel_rate)) +
  geom_col(fill = "tomato") +
  labs(
    title = "Tasa de cancelación por país (Top 10)",
    x = "País",
    y = "Proporción cancelada"
  ) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

```

Cancellation per type (tipo) of trip

```{r dif_cancellation_tipo}

xx <- x %>%
  mutate(is_canceled = as.numeric(as.character(is_canceled)))

cancel_by_tipo <- xx %>%
  group_by(tipo) %>%
  summarise(
    cancel_rate = mean(is_canceled),
    total = n()
  )

ggplot(cancel_by_tipo, aes(x = reorder(tipo, -cancel_rate), y = cancel_rate)) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Tasa de cancelación por tipo de viaje",
    x = "Tipo de viaje",
    y = "Proporción cancelada"
  ) +
  scale_y_continuous(labels = scales::percent) +
  theme_minimal()

```

Country and trip type – Combined cancellation rate

```{r combined_cancellation}

xx <- x %>%
  mutate(is_canceled = as.numeric(as.character(is_canceled)))

cancel_heatmap <- xx %>%
  filter(country %in% top_countries) %>%
  group_by(country, tipo) %>%
  summarise(cancel_rate = mean(is_canceled), .groups = "drop")

ggplot(cancel_heatmap, aes(x = tipo, y = country, fill = cancel_rate)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(labels = scales::percent) +
  labs(
    title = "Tasa de cancelación por país y tipo de viaje",
    x = "Tipo de viaje",
    y = "País",
    fill = "Cancelado"
  ) +
  theme_minimal()

```

**3)** relationship between type of stay 'tipo' and cost 'adr'

Boxplot: Compare 'adr' distribution by 'tipo'
Ideal for viewing medians, ranges, and price outliers.

```{r boxplot_adr_tipo}

ggplot(x, aes(x = tipo, y = adr, fill = tipo)) +
  geom_boxplot() +
  labs(
    title = "Distribution of daily cost (adr) by type of stay",
    x = "Type of stay",
    y = "ADR (Average Daily Rate)"
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€"))

```

Dot plot with mean and standard error
Shows the mean ADR by type, with error bars.

```{r dot_plot_adr_tipo}

xx %>%
  group_by(tipo) %>%
  summarise(
    mean_adr = mean(adr, na.rm = TRUE),
    se = sd(adr, na.rm = TRUE) / sqrt(n())
  ) %>%
  ggplot(aes(x = reorder(tipo, -mean_adr), y = mean_adr)) +
  geom_col(fill = "skyblue") +
  geom_errorbar(aes(ymin = mean_adr - se, ymax = mean_adr + se), width = 0.2) +
  labs(
    title = "Mean cost per type of stay",
    x = "Type of stay",
    y = "ADR medio (€)"
  ) +
  theme_minimal() +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€"))

```

Violin plot

```{r violin_plot_adr_tipo}

ggplot(x, aes(x = tipo, y = adr, fill = tipo)) +
  geom_violin(trim = FALSE) +
  geom_boxplot(width = 0.1, fill = "white", outlier.size = 0.5) +
  labs(
    title = "ADR distribution per type of stay",
    x = "Type of stay",
    y = "ADR (€)"
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€"))


```

**4)** differences among groups with respect to hotel type (city / resort)

Stacked Bar Plot: Proportions of Stay Types by Hotel Type
This graph shows the distribution of stay types (tipo) within each hotel type, with bars representing the proportion of each stay type.

```{r stackedbar_plot_hotel_tipo}

xx %>%
  count(hotel, tipo) %>%
  group_by(hotel) %>%
  mutate(proportion = n / sum(n)) %>%
  ggplot(aes(x = hotel, y = proportion, fill = tipo)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Proportions of Stay Types by Hotel Type",
    x = "Hotel Type",
    y = "Proportion of Stay Type"
  ) +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent) +
  theme(legend.position = "bottom")

```

Boxplot for ADR by Hotel Type and Stay Type
This will show how adr varies between the two hotel types (Resort Hotel and City Hotel) for each tipo (stay type).

```{r boxplot_hotel_tipo}

ggplot(x, aes(x = tipo, y = adr, fill = hotel)) +
  geom_boxplot() +
  facet_wrap(~ hotel) +  # Separate boxplots by hotel type
  labs(
    title = "Distribution of ADR by Stay Type and Hotel Type",
    x = "Tipo de estancia",
    y = "ADR (Average Daily Rate)"
  ) +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_y_continuous(labels = scales::dollar_format(prefix = "€"))

```


</font>

NOTE: This is a good example of using ChatGPT or other generative AI to ask 
interesting questions about the proposed dataset. The following paper describes
the potential uses of generative AI in the different phases of creating a data
visualization for storytelling:

https://ieeexplore.ieee.org/stamp/stamp.jsp?arnumber=10891192


Delete unnecessary columns to reduce dataset size:

```{r reduce_dataset}

x_reduced <- x %>% select(-company, -agent, -reserved_room_type, -assigned_room_type, -market_segment, -distribution_channel, -arrival_date_week_number, -customer_type)

write.csv(x_reduced, "x_reduced.csv", row.names = FALSE, quote = TRUE)

```
